\documentclass[runningheads]{llncs}
\makeatletter
\def\input@path{{sections/}{figures/}{bib/}}
\makeatother
\raggedbottom

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{svg}
\usepackage{booktabs}
\usepackage{xurl}
<<<<<<< HEAD
\usepackage{hyperref} % removes red boxes around Figure refs/links

% \usepackage{subfig}
=======
\usepackage[justification=centering]{caption}
\usepackage{float}
\usepackage{multirow}
\usepackage{placeins}
\usepackage{hyperref}
>>>>>>> main

% ---------- Double-blind toggle ----------
\newif\ifanonymized
\anonymizedtrue % review submission: true ; camera-ready: false

% Make figures resolve without ../
\graphicspath{{figures/}}

\begin{document}

% ---------- Title ----------
\title{MitoChime: A Machine Learning Pipeline for Detecting PCR-Induced Chimeras in Mitochondrial Illumina Reads}
\titlerunning{MitoChime: Chimera Detection in mtDNA Reads}

% ---------- Author block ----------
\ifanonymized
  \author{X X X X X}
  \authorrunning{X}
  \institute{X}
\else
  \author{Yvonne Lin\inst{1} \and Duranne Duran\inst{1} \and Daniella Pailden\inst{1}}
  \authorrunning{Y. Lin et al.}
  \institute{University of the Philippines Visayas, Miagao, Iloilo, Philippines\\
  \email{linyvonnea@gmail.com}}
\fi

\maketitle



% ---------- Abstract ----------
\begin{abstract}
Polymerase chain reaction (PCR) amplification during Illumina library preparation can introduce chimeric reads that degrade mitochondrial genome assembly by creating spurious junctions and fragmented assembly graphs. Existing chimera detection tools are largely optimized for amplicon-based microbial workflows and often rely on reference databases or abundance assumptions that do not fit single-species organellar sequencing. We present a pipeline for detecting PCR-induced chimeric reads in \textit{Sardinella lemuru} mitochondrial paired-end data using alignment-derived features and a sequence-based convolutional model. Using simulated clean and chimeric reads, alignment-structure and clipping metrics are extracted alongside k-mer discontinuity and microhomology descriptors as features. On a held-out test set, a tuned Gradient Boosting classifier achieves F1=0.777 and ROC--AUC=0.846 for chimeric-read detection, while a 1D Convolutional Neural Network model achieves F1=0.914 and ROC--AUC=0.961 with higher recall. Assembly evaluation using SPAdes-generated assembly graphs summarized with Bandage statistics shows that 
filtering reduced assembly graph complexity at high chimera rates, with performance depending on read depth. These results support MitoChime as a practical pre-assembly quality-control step for mitochondrial genome assembly pipelines.
\keywords{Chimera detection \and Mitochondrial genome assembly \and Illumina reads \and Machine learning \and Deep learning}
\end{abstract}

% from documentation/nids2026/src/main.tex
\input{sections/01_introduction.tex}
\input{sections/02_related_work.tex}
\input{sections/03_methods.tex}
\input{sections/04_results.tex}
\input{sections/05_discussion.tex}
\input{sections/06_conclusion.tex}

\bibliographystyle{splncs04}
\bibliography{bib/references}


\end{document}