%   Filename    : chapter_2.tex

\chapter{Review of Related Literature}
\label{sec:relatedlit}

This chapter presents an overview of the literature relevant to the study. It discusses the biological and methodological foundations underlying mitochondrial genome analysis and assembly, PCR chimera formation, as well as existing tools, algorithms, and techniques related to chimera detection. The chapter aims to examine and highlight the strengths, limitations, and research gaps in current approaches that motivate the development of the present study.

\section{The Mitochondrial Genome}

The mitochondrial genome, or mtDNA, is a small and usually circular molecule found in most eukaryotic organisms. It encodes genes involved in oxidative phosphorylation and cellular energy metabolism. Because of its conserved structure mtDNA has become a valuable genetic marker for studies in population genetics and phylogenetics \citep{Anderson1981,Boore1999}. In animal species, the mitochondrial genome is typically around 15--20 kilobases long and contains 13 protein-coding genes, 22 transfer RNA genes, and two ribosomal RNA genes arranged compactly without introns \citep{Gray1999}. In comparison to nuclear DNA, the ratio of the number of copies of mtDNA is higher and has simple organization which make it particularly suitable for genome sequencing and assembly studies \citep{Dierckxsens2017}.

\subsection{Mitochondrial Genome Assembly}

Mitochondrial genome assembly refers to the reconstruction of the complete mitochondrial DNA (mtDNA) sequence from raw or fragmented sequencing reads. It is conducted to obtain high-quality, continuous representations of the mitochondrial genome that can be used for a wide range of analyses, including species identification, phylogenetic reconstruction, evolutionary studies, and investigations of mitochondrial diseases. Because mtDNA evolves rapidly, its assembled sequence provides valuable insights into population structure, lineage divergence, and adaptive evolution across taxa \citep{Boore1999}. Compared to nuclear genome assembly, assembling the mitochondrial genome is often considered more straightforward but still encounters technical challenges such as the formation of chimeric reads. Commonly used tools for mitogenome assembly such as SPAdes, GetOrganelle and MITObim operate under the assumption of organelle genome circularity, and are vulnerable when chimeric reads disrupt this circular structure, resulting in assembly errors \citep{Jin2020,Hahn2013}.

\section{PCR Amplification and Chimera Formation}

PCR plays an important role in NGS library preparation, as it amplifies target DNA fragments for downstream analysis. However as previously mentioned, the amplification process can also introduce chimeric reads which compromises the quality of the input reads supplied to sequencing or assembly workflows. Chimeras typically arise when incomplete extension occurs during a PCR cycle. This causes the DNA polymerase to switch from one template to another and generate hybrid recombinant molecules \citep{Judo1998}. Artificial chimeras are produced through such amplification errors, whereas biological chimeras occur naturally through genomic rearrangements or transcriptional events.

In the context of amplicon-based sequencing, the presence of chimeras can inflate estimates of genetic or microbial diversity and may cause misassemblies during genome reconstruction. \cite{Qin2023} has reported that chimeric sequences may account for more than 10\% of raw reads in amplicon datasets. This artifact tends to be most prominent among rare operational taxonomic units (OTUs) or singletons, which are sometimes misinterpreted as novel diversity, further causing the complication of microbial diversity analyses \citep{Gonzalez2004}. As such, determining and minimizing PCR-induced chimera formation is vital for improving the quality of mitochondrial genome assemblies, and ensuring the reliability of amplicon sequencing data.

\section{Existing Traditional Approaches for Chimera Detection}
\label{sec:abundance_skew}

Computational approaches for chimera detection have generally been grouped into two broad categories: reference-based detection and de novo detection. Reference-based chimera detection, also known as database-dependent detection, is one of the earliest and most widely used computational strategies for identifying chimeric sequences in amplicon-based community studies. These methods rely on the comparison of each query sequence against a curated, high-quality database of known, non-chimeric reference sequences \citep{Edgar2011}.

On the other hand, de novo chimera detection, also referred to as reference-free detection, represents an alternative computational paradigm that identifies chimeric sequences without reliance on external reference databases. This method infers chimeras based on internal relationships among the sequences present within the dataset itself, making it particularly advantageous in studies of underexplored or taxonomically diverse communities where comprehensive reference databases are unavailable or incomplete \citep{Edgar2011,Edgar2016}. The underlying assumption in this method is that during PCR, true biological sequences are generally more abundant as they are amplified early and dominate the read pool, whereas chimeric sequences appear later and are generally less abundant. The de novo approach leverages this abundance hierarchy, treating the most abundant sequences as supposed parents and testing whether less abundant sequences can be reconstructed as mosaics of these templates. Compositional and structural similarity are also evaluated to check whether different regions of a candidate sequence correspond to distinct high-abundance sequences.

In practice, many modern bioinformatics pipelines combine both paradigms sequentially: an initial de novo step identifies dataset-specific chimeras, followed by a reference-based pass that removes remaining artifacts relative to established databases \citep{Edgar2016}. These two methods of detection form the foundation of tools such as UCHIME and later UCHIME2.

\subsection{UCHIME}

UCHIME is one of the most widely used tools for detecting chimeric sequences in amplicon-based studies and remains a standard quality-control step in microbial community analysis. Its core strategy is to test whether a query sequence (\(Q\)) can be explained as a mosaic of two parent sequences, (\(A\) and \(B\)), and to score this relationship using a structured alignment model \citep{Edgar2011}.

In reference mode, UCHIME divides the query into several segments and maps them against a curated database of non-chimeric sequences. Candidate parents are identified, and a three-way alignment is constructed. The algorithm assigns “Yes” votes when different segments of the query match different parents and “No” votes when the alignment contradicts a chimeric pattern. The final score reflects the balance of these votes. In de novo mode, UCHIME operationalizes the abundance-skew principle described earlier: high-abundance sequences are treated as candidate parents, and lower-abundance sequences are evaluated as potential mosaics. This makes the method especially useful when no reliable reference database exists.

Although UCHIME is highly sensitive, it faces key constraints. Chimeras formed from parents with very low divergence (below \(\sim 0.8\%\)) are difficult to detect because they are nearly indistinguishable from sequencing errors. Accuracy in reference mode depends strongly on database completeness, while de novo detection assumes that true parents are both present and sufficiently more abundant, conditions that are not always met.

\subsection{UCHIME2}

UCHIME2 extends the original algorithm with refinements tailored for high-resolution sequencing data. One of its major contributions is a re-evaluation of benchmarking practices. \cite{Edgar2016} demonstrated that earlier accuracy estimates for chimera detection were overly optimistic because they relied on unrealistic scenarios where all true parent sequences were assumed to be present. Using the more rigorous CHSIMA benchmark, UCHIME2 showed the prevalence of “fake models” or real biological sequences that can be perfectly reconstructed as apparent chimeras of other sequences, which suggests that perfect chimera detection is theoretically unattainable. UCHIME2 also introduces several preset modes (e.g., denoised, balanced, sensitive, specific, high-confidence) designed to tune sensitivity and specificity depending on dataset characteristics. These modes allow users to adjust the algorithm to the expected noise level or analytical goals.

Despite these improvements, UCHIME2 must be applied with caution. The website manual explicitly advises against using UCHIME2 as a standalone chimera-filtering step in OTU clustering or denoising workflows because doing so can inflate both false positives and false negatives \citep{EdgarManual}.

\subsection{DADA2}

Building upon the principles of de novo chimera detection, Callahan et al. (2016) developed the Divisive Amplicon Denoising Algorithm 2 (DADA2), a model-based approach that distinguishes true biological sequence variation from sequencing error at single-nucleotide resolution. DADA2 is an inference pipeline developed for Illumina data, particularly 16S rRNA gene sequencing. Unlike OTU-based methods, it infers exact amplicon sequence variants (ASVs) through quality filtering, dereplication, denoising, paired-end merging, and chimera removal \citep{Callahan2016}.

Its chimera detection step is implemented through \texttt{isBimeraDenovo()}, which identifies sequences that can be reconstructed as exact bimeras, or two-parent chimeras, from more abundant inferred sequences. The method performs Needleman--Wunsch global alignment between a candidate sequence and more abundant sequences, then searches for a left-parent and right-parent combination that can explain the child sequence. This step is applied after denoising and is intended for exactly inferred sample sequences rather than noisy input reads or fuzzy OTUs \citep{Callahan2016}.

Although DADA2 is effective for amplicon-based microbial studies, it is not designed for read-level detection of PCR-induced chimeras in single-species organellar sequencing data. Its dependence on ASV-level reconstruction and abundance relationships limits its direct applicability to the focus of the present study.

\subsection{CATCh}

Early approaches to chimera detection such as UCHIME relied on alignment-based sequences in amplicon data. However, researchers soon observed that different algorithms often produced inconsistent predictions. A sequence might be identified as chimeric by one tool but classified as non-chimeric by another, resulting in unreliable filtering outcomes across studies.

To address these inconsistencies, \cite{Mysara2015} developed the Classifier for Amplicon Tool Chimeras (CATCh), which represents the first ensemble machine learning system designed for chimera detection in 16S rRNA amplicon sequencing. Rather than depending on a single detection strategy, CATCh integrates the outputs of several established tools, including UCHIME, ChimeraSlayer, DECIPHER, Pintail, and Perseus. The individual scores and binary decisions generated by these tools are used as input features for a supervised learning model. The algorithm employs a Support Vector Machine (SVM) with a Pearson VII Universal Kernel (PUK) to determine optimal weightings among the input features and to assign each sequence a probability of being chimeric.

Benchmarking in both reference-based and de novo modes demonstrated significant performance improvements. CATCh achieved sensitivities of approximately 85 percent in reference-based mode and 92 percent in de novo mode, with corresponding specificities of approximately 96 percent and 95 percent. These results indicate that CATCh detected 7 to 12 percent more chimeras than any individual algorithm while maintaining high precision.

\subsection{ChimPipe}

Among the available tools for chimera detection, ChimPipe is a pipeline developed to identify chimeric sequences such as biological chimeras. It uses both discordant paired-end reads and split-read alignments to improve the accuracy and sensitivity of detecting biological chimeras \citep{Rodriguez2017}. By combining these two sources of information, ChimPipe achieves better precision than methods that depend on a single type of indicator.

The pipeline works with many eukaryotic species that have available genome and annotation data \citep{Rodriguez2017}. It can also predict multiple isoforms for each gene pair and identify breakpoint coordinates that are useful for reconstructing and verifying chimeric transcripts. Tests using both simulated and real datasets have shown that ChimPipe maintains high accuracy and reliable performance.

ChimPipe lets users adjust parameters to fit different sequencing protocols or organism characteristics. Experimental results have confirmed that many chimeric transcripts detected by the tool correspond to functional fusion proteins, demonstrating its utility for understanding chimera biology and its potential applications in disease research \citep{Rodriguez2017}.

\section{Machine Learning and Deep Learning Approaches for Chimera and Sequence Quality Detection}

Traditional chimera detection tools rely primarily on heuristic or alignment-based rules. Recent advances in machine learning (ML) have demonstrated that models trained on sequence-derived features can effectively capture compositional and structural patterns in biological sequences. Although most existing ML systems such as those used for antibiotic resistance prediction, taxonomic classification, or viral identification are not specifically designed for chimera detection, they highlight how data-driven models can outperform similarity-based heuristics by learning intrinsic sequence signatures \citep{Arango2018,Liang2020,Ren2020}.

In this broader literature, two major modeling directions are especially relevant. The first is feature-based machine learning, in which reads are represented through engineered descriptors such as k-mer frequencies, GC-related composition, and alignment-derived statistics. The second is sequence-based deep learning, in which models learn directly from encoded nucleotide input and can capture local motifs as well as longer-range dependencies without requiring all relevant patterns to be defined manually in advance.

\subsection{Feature-Based Representations of Genomic Sequences}

Feature extraction converts DNA sequences into numerical representations suitable for machine learning models. One approach is k-mer frequency analysis, which counts short nucleotide sequences within a read \citep{Vervier2015}. High-frequency k-mers, including simple repeats such as “AAAAAA,” can highlight repetitive or unusual regions that may occur near chimeric junctions. Comparing k-mer patterns across adjacent parts of a read can help identify such regions, while GC content provides an additional descriptor of local sequence composition \citep{Ren2020}.

Alignment-derived features further inform junction detection. Long-read tools such as Sniffles \citep{Sedlazeck2018} use split alignments to locate breakpoints across extended sequences, whereas short-read aligners like Minimap2 \citep{Li2018} report supplementary and secondary alignments that indicate local discontinuities. Split alignments, where parts of a read map to different regions, can reveal template-switching events. These features complement k-mer profiles and enhance detection of potentially chimeric reads, even in datasets with incomplete references.

Microhomology, or short sequences shared between adjacent segments, is another biologically meaningful feature. Short microhomologies, typically 3--20 bp, are involved in template switching both in cellular repair pathways and during PCR, where they act as signatures of chimera formation \citep{Sfeir2015,Peccoud2018}. In PCR-induced chimeras, short identical sequences at junctions provide a clear signature of chimerism. Measuring the longest exact overlap at each breakpoint complements k-mer and alignment features and helps identify reads that are potentially chimeric.

\subsection{Sequence-Based Deep Learning Representations of Genomic Sequences}

For deep learning models to process biological data, raw DNA or RNA sequences must be mapped into a numerical coordinate system. \citet{He2021} categorized several common preprocessing techniques for motif mining.

One standard approach is one-hot encoding, where each nucleotide is transformed into a sparse binary vector, such as \(A = [1, 0, 0, 0]\). This method preserves both the discrete identity of the bases and their sequence order, making it suitable for convolutional models that scan across nucleotide positions. Another common method is k-mer tokenization. In this approach, a sequence is divided into overlapping subsequences of length \(k\), and each k-mer is treated as a token. These tokens can then be converted into integer indices and passed through an embedding layer. This allows the model to learn dense vector representations of recurring sequence patterns. In the present study, these encoding paradigms served as the foundation for extracting features from the short-read sequences.

\subsection{CNN and RNN for Genomic Sequence Analysis}

Convolutional neural networks (CNNs) have been widely used in genomic sequence analysis because they can learn local sequence patterns directly from encoded DNA or RNA inputs. In these models, convolutional filters slide across the sequence to detect recurring motifs, splice signals, binding sites, and short-range changes in nucleotide composition.

A foundational implementation of this approach is DeepBind \citep{Alipanahi2015}, which used 1D CNNs to identify the binding specificities of DNA- and RNA-binding proteins. In this framework, each convolutional filter acts as a non-linear motif detector, learning to recognize specific binding sites directly from the raw sequence data. Building upon this motif-recognition capability, Deep Splicer \citep{FernandezCastillo2022} employed a CNN architecture to identify splice sites across diverse eukaryotic genomes. The study found that the model significantly reduced false-positive predictions compared to traditional tools and demonstrated a high degree of generalizability across species, including humans and mice. Beyond classification, CNNs can also be used for quantitative modeling. The DeFine framework \citep{Wang2018} used deep convolutional layers to accurately quantify transcription factor-DNA binding intensities and evaluate functional non-coding variants.

In the context of PCR-induced chimera detection, using CNNs is relevant because chimeric reads may contain local breakpoint patterns, unusual sequence transitions, or short regions that differ from clean read structures. CNN can scan across the nucleotide sequence and learn filters that respond to these local signals.

Recurrent neural networks (RNNs) are designed to model sequential data. In genomic sequence analysis, this makes them useful for tasks where the arrangement of nucleotides or k-mers affects the prediction. Unlike CNNs, which primarily capture local patterns through convolutional filters, recurrent models can capture features across positions in a sequence.

\citet{Shen2018} utilized this approach in developing KEGRU, a model designed to identify transcription factor binding sites in DNA sequences by combining k-mer embedding with a Bidirectional Gated Recurrent Unit (BiGRU) network. The results of the study showed better performance compared to other models that used CNN-based feature extraction.

Beyond standalone deep learning architectures, hybrid models that leverage the strengths of both CNNs and RNNs have demonstrated strong performance in complex sequence classification tasks. \citet{Tabane2025} demonstrated this through a hybrid CNN-BiLSTM model, where the convolutional layers served as automated feature extractors that identified local patterns, which were then fed into the Bidirectional Long Short-Term Memory (BiLSTM) layers to capture long-distance dependencies.

Overall, these studies show that CNNs and RNNs are well suited for genomic sequence analysis. Although direct deep learning research or tools focused specifically on PCR-induced chimera detection in the mitochondrial genome remain limited, the previously presented studies show that neural networks can extract biologically meaningful patterns from various genomic data.

\section{Synthesis of Existing Tools and Machine Learning Approaches}

To provide an integrated overview of the literature discussed in this chapter, Table~\ref{tab:study_summary} summarizes the existing chimera detection tools, traditional machine learning approaches, and deep learning-based genomic sequence models reviewed in this study. The synthesis highlights each method or tool’s core approach and key limitations in relation to the present study.

\newgeometry{top=50pt, bottom=60pt, left=30pt, right=30pt}
\begin{landscape}

\begin{xltabular}{\linewidth}{>{\bfseries}p{3.2cm} X X}
\caption{Comparison of Chimera Detection Tools and Genomic Sequence Modeling Approaches}
\label{tab:study_summary} \\

\toprule
Method / Tool & \textbf{Core Approach} & \textbf{Key Limitations} \\
\midrule
\endfirsthead

\caption[]{Comparison of Chimera Detection Tools and Genomic Sequence Modeling Approaches (continued)} \\

\toprule
Method / Tool & \textbf{Core Approach} & \textbf{Key Limitations} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\textit{Continued on the next page}} \\
\endfoot

\bottomrule
\endlastfoot

Reference-based Detection &
Compares each query sequence against curated databases of verified, non-chimeric sequences; evaluates segment similarity to identify mosaic patterns. &
Accuracy depends on database completeness; performs poorly for novel taxa or missing parents; limited sensitivity for low-divergence chimeras. \\
\addlinespace

De novo Detection &
Identifies chimeras using only internal dataset structure; leverages abundance hierarchy and compositional similarity to infer whether low-abundance sequences can be reconstructed from abundant parents. &
Assumes true sequences are more abundant; fails when amplification bias distorts abundances; struggles when parental sequences are similarly abundant or highly similar. \\
\addlinespace

UCHIME &
Alignment-based model that partitions the query into segments, identifies parent candidates, and computes a chimera score via a three-way alignment; supports reference and de novo modes. &
Reduced accuracy for very closely related parents (\(<0.8\%\) divergence); sensitive to incomplete databases; de novo mode fails if parents are absent or not sufficiently more abundant. \\
\addlinespace

UCHIME2 &
Updated UCHIME with improved benchmarking (CHSIMA) and multiple sensitivity/specificity presets; better handles incomplete references and dataset variability. &
“Fake models” limit theoretical accuracy; genuine variants may mimic chimeras; not recommended as a standalone step in OTU or denoising pipelines due to increased false positives/negatives. \\
\addlinespace

DADA2 &
Infers exact amplicon sequence variants from Illumina reads through quality filtering, dereplication, denoising, paired-end merging, and chimera removal; uses \texttt{isBimeraDenovo()} to identify exact two-parent bimeras from more abundant inferred sequences. &
Designed for ASV-level amplicon analysis, especially 16S rRNA data; not intended for direct read-level detection of PCR-induced chimeras in single-species mitochondrial sequencing data. \\
\addlinespace

CATCh &
First ensemble ML model for 16S chimera detection; integrates outputs of UCHIME, ChimeraSlayer, DECIPHER, Pintail, and Perseus using an SVM to boost overall prediction accuracy. &
Performance constrained by underlying tools; ML model cannot capture features not present in component algorithms; may misclassify in highly novel or low-coverage datasets. \\
\addlinespace

ChimPipe &
Pipeline for detecting biological chimeras in RNA-seq using discordant paired-end reads and split-read alignments; identifies isoforms and breakpoint coordinates. &
Requires high-quality genome and annotation; tailored to RNA-seq rather than amplicons; computationally intensive; limited to organisms with available reference genomes. \\
\addlinespace

CNN Approach &
Applies one-dimensional convolutional filters to encoded DNA or RNA sequences to learn local motifs, splice signals, binding patterns, and short-range sequence transitions. &
Requires careful encoding; learned features may be difficult to interpret; direct use for mitochondrial PCR-induced chimera detection remains limited. \\
\addlinespace

RNN Approach &
Treats nucleotides or k-mers as ordered sequences and uses recurrent layers to capture positional and longer-range dependencies across genomic reads. &
More computationally demanding than simpler ML models; performance depends on representation and tuning; direct use for mitochondrial PCR-induced chimera detection remains limited. \\
\addlinespace

Hybrid CNN--RNN Models &
Combines CNN-based local feature extraction with RNN-based sequence modeling to capture both short-range genomic patterns and longer-distance dependencies. &
More complex to train and tune; requires sufficient labeled data; existing studies support genomic sequence classification generally, but not specifically PCR-induced mitochondrial chimera detection. \\

\end{xltabular}

\end{landscape}
\restoregeometry

Across existing studies, no single approach reliably detects all forms of chimeric sequences, and the reviewed literature consistently shows that chimeras remain a persistent challenge in genomics and bioinformatics. Although the surveyed tools are not designed specifically for organelle genome assembly, they provide valuable insights into which methodological strategies are effective and where current approaches fall short. These limitations collectively define a clear research gap: the need for a specialized, feature-driven detection framework tailored to PCR-induced mitochondrial chimeras. Addressing this gap aligns with the research objective outlined in Chapter~1, which is to develop and evaluate a machine learning-based pipeline (MitoChime) that improves the quality of downstream mitochondrial genome assembly. In support of this aim, the subsequent chapters describe the design, implementation, and evaluation of the proposed tool.