%   Filename    : chapter_5.tex

\chapter{Conclusion}

\section{Summary of Findings}

This study evaluated machine learning and deep learning approaches for detecting PCR-induced chimeric reads in simulated mitochondrial Illumina data. Across all experiments, the results showed that both engineered-feature classifiers and sequence-based neural models can distinguish clean from chimeric reads, but their performance differed substantially in predictive accuracy, feature dependence, and downstream assembly effects.

Among the engineered-feature approaches, all meaningful classifiers outperformed the dummy baseline, confirming that the extracted features captured informative signals related to chimeric read formation. Baseline and tuned experiments showed that tree-based ensemble methods consistently ranked highest among the classical models. In particular, Gradient Boosting achieved the best held-out test performance among the feature-based classifiers, with an F1-score of 0.7765 and a ROC--AUC of 0.8459, while maintaining the shortest runtime of 0.50 minutes. These results indicate that engineered alignment-derived and sequence-derived features are sufficient for reliable chimera detection, although their performance remained below that of the deep learning models.

Feature analysis further clarified which engineered signals were most useful. Descriptive statistics, boxplots, and permutation importance consistently showed that clipping-based features, especially \texttt{total\_clipped\_bases}, provided the strongest discrimination between clean and chimeric reads. K-mer divergence features also contributed meaningful predictive signal, although their standalone class separation was weaker than that of clipping features. In contrast, microhomology-related variables and several supplementary-alignment structure features contributed little to overall predictive performance. This pattern was reinforced by the feature selection results, where a reduced four-feature subset achieved nearly identical performance to the full engineered-feature model, and removal of explicit microhomology features produced only negligible loss in classification quality.

The deep learning models substantially outperformed the feature-based classifiers. The one-dimensional CNN achieved a held-out test F1-score of 0.9135 and a ROC--AUC of 0.9607, demonstrating that direct sequence-based modeling can capture chimera-related patterns beyond those represented by the engineered features alone. The RNN-kmer BiGRU achieved the best overall classification performance in the study, with a held-out test accuracy of 0.9516, precision of 0.9394, recall of 0.9655, F1-score of 0.9523, and ROC--AUC of 0.9849. Its cross-validation performance was also stable across folds, indicating strong generalization under the simulated setting. The three top-performing models also showed differences in performance. Gradient Boosting was the strongest among the classical models, the CNN substantially outperformed the feature-based approaches, and the RNN-kmer BiGRU achieved the highest overall predictive performance. This improvement, however, required more computation time, with Kaggle runtimes of 5.51 minutes for the BiGRU, 1.39 minutes for the CNN, and 0.50 minutes for Gradient Boosting.

Beyond read-level classification, external assembly testing showed that model choice also affected downstream assembly structure. At low coverage, assembly outcomes depended strongly on read retention. Gradient Boosting was the most reliable under these conditions because it consistently retained the largest proportion of reads while generally preserving contiguous assemblies. BiGRU showed intermediate behaviour: it removed more reads than Gradient Boosting and reduced graph complexity more strongly, but under high chimera rates at the lowest read depth, this reduction in read support could also fragment the assembly. CNN applied the strongest filtering, which at low depth sometimes caused fragmentation through aggressive read removal. At higher coverage, however, stronger filtering became more advantageous. In the 20K paired-end reads, 50\% chimeras, CNN produced the simplest assembly graph, while BiGRU achieved only partial simplification and Gradient Boosting retained more residual graph structure. These findings show that the model with the best read-level classification metrics is not always the model that yields the simplest downstream assembly graph, because assembly behaviour depends not only on how many reads are removed but also on which reads are removed.

Taken together, the results show that chimera detection in simulated mitochondrial reads can be performed effectively using both engineered-feature and deep learning methods, but sequence-based neural models provide a performance advantage. At the same time, the downstream assembly experiments highlight that predictive performance alone is not sufficient for judging practical utility. Instead, chimera filtering methods should be evaluated jointly in terms of classification quality, computational cost, read retention, and assembly consequences.

\section{Conclusion}

This study demonstrates that PCR-induced chimeric reads in simulated mitochondrial Illumina data can be identified effectively before assembly using both feature-based machine learning and sequence-based deep learning models. Among the engineered-feature approaches, Gradient Boosting achieved the strongest performance and indicated that clipping-related and k-mer divergence features contributed most of the useful predictive information. Across all evaluated models, however, the RNN-kmer BiGRU achieved the highest overall classification performance, showing that sequential k-mer representations are particularly effective for this detection task.

The findings further indicate that the most appropriate model depends on the intended application. The RNN-kmer BiGRU was the strongest model for read-level prediction, Gradient Boosting was the most practical feature-based alternative in terms of runtime and interpretability, and the CNN showed the strongest graph simplification in the high-coverage, high-chimera external assembly condition. Accordingly, automated chimera detection is supported as a useful pre-assembly filtering step, but model choice should consider downstream assembly effects in addition to predictive performance.

\section{Limitations of the Study}

Several limitations should be considered when interpreting the results. First, all experiments were conducted on simulated mitochondrial Illumina reads. Although this provides controlled ground truth for evaluating chimera detection, real sequencing data may exhibit additional noise sources, alignment irregularities, and biological complexities that are not fully captured in simulation. Second, the dataset was balanced by design, whereas real sequencing workflows may contain very different class proportions. Third, the study focused on a specific set of engineered features and two sequence-based neural architectures; other architectures, tokenization strategies, and hybrid feature-sequence models were not explored. Fourth, assembly validation was limited to selected external test conditions and primarily interpreted through graph structure and read retention, rather than broader downstream biological analyses.

\section{Recommendations for Future Work}

Future work should extend the evaluation to real mitochondrial and non-mitochondrial sequencing datasets to determine how well the observed performance generalizes under practical laboratory conditions. Additional experiments should also examine class imbalance, broader chimera-generation mechanisms, and sequencing-error patterns that better reflect empirical data. On the modeling side, hybrid approaches that combine engineered alignment features with deep sequence encoders may further improve robustness and interpretability.