%   Filename    : chapter_1.tex 
    \section*{Chapter 1 }
\label{sec:researchdesc}    %labels help you reference sections of your document
\section{Introduction}
\subsection{Overview}
\label{subsec:overview}
    
    Philippine folk literature is the body of oral literature of the Filipino people. Folk literature typically undergoes classification into three categories: folk narratives, folk speech, and folk songs. Myths, legends, and folktales are included in the category of folk narratives, a form of literature that provides a narrative through prose or verse, and will serve as the focus of this project. Myths and legends are both regarded as truthful accounts of the past that provide explanations for the origins of entities in the environment. However, myths are often sacred and linked with religion, whereas legends tend to be secular in nature. On the other hand, folktales are fictitious prose narratives typically employed for entertainment purposes \cite{eugenio2007philippine}. In addition to their roles in explaining origins or providing entertainment, these three forms of folk literature often function as mediums for the communication of morals, traditions, and beliefs of the Filipino people. \shortciteA{eslit2023resilience} explored 10 popular folklores in the Philippines, examining their portrayal of Filipino culture and identity. Common themes in the analyzed folklore include environmental importance, respect for elders, and justice. These forms of folk literature have played significant roles in the conveyance and instillment of key values, traditions, and identity within particular ethnolinguistic groups. However, as \citeA{eugenio2007philippine} notes, there is a significant lack of collections of Philippine folk literature. Consequently, research on Philippine folk literature presents difficulties due to its wide dispersion across the country, the necessity for translations, and the rapid decline of this literary form, which limits available research. While there has been some work addressing these challenges, access has been limited due to cost and dated nature.

    According to \cite{dimzon2015}, there exists no digital ontology of Western Visayas folklore as digital ontology development was a new area of research. Their pioneering work serves as the start of the digitization of the Western Visayas folklore and is the basis of the researchers’ work. 
    With this, researchers propose the development of an ontology-based chatbot capable of answering questions and providing information about folk narratives, particularly those from Panay. 
    
     \shortciteA{jepsen2009just} offers a practical definition of ontology. Specifically, ontology as "a method of representing items of knowledge (ideas, facts, things—whatever) in a way that defines the relationships and classifications of concepts within a specified domain of knowledge." A chatbot is a software agent with the capability for engaging in human-like conversation. The researchers aim to provide the chatbot with knowledge and understanding of the relationships between concepts found in Panayanon folk narratives, which enables it to answer queries about them. Through the proposed system, the creation of a central hub of knowledge on Panayanon folk narratives facilitates the streamlining and accessibility of research and education on Panayanon folk narratives. Furthermore, the proposed system contributes to the preservation and promotion of cultural diversity and heritage, as globalization heightens the threat of the deterioration and disappearance of cultural heritage \cite{UNESCO_2001}. 

\subsection{Problem Statement}
\label{subsec:probstatement}
The body of knowledge regarding Philippine cultural heritage, specifically Philippine folk literature, remains limited. Despite efforts to collect and analyze this literature, the accessibility of such research is constrained by the cost of resources and the outdated nature of existing works. \citeA{eugenio2007philippine} affirms the lack of comprehensive collections and accessible resources on Philippine folk literature, resulting in significant challenges in the study, documentation, and promotion of this literary form.  

Damiana Eugenio, recognized as "Ina ng Folklor ng Pilipinas" by the U.P. Folklorists, Inc. and the U.P. Folklore Studies Program, has made significant contributions to the preservation of Philippine cultural heritage. Her book Philippine Folk Literature: An Anthology—the first volume in a seven-volume series—compiled over 150 texts and selections of proverbs and riddles from across the Philippines. However, due to the rapid digitization of global information and the fact that her works are now over 15 years old, their accessibility continues to diminish.

Recent efforts have sought to address this issue, with projects like the Aswang Project, created in 2006 by Jordan Clark. This project serves as an online resource for Philippine folklore, featuring articles about various myths, creatures, and spirits found throughout the country. Furthermore, in the terminal report of \citeA{dimzon2015}, they have collected and digitized Panayanon myths and legends by creating ontologies using Web Ontology Language (OWL). However, their work is not made publicly available and has not included folk tales from Panay; gaps remain in the collection of Panayanon folk narratives, which the researchers aim to explore further.

In the field of chatbots, \citeA{shawar2007chatbots} note that chatbots are designed to accommodate users' natural tendency to express their wishes through speaking, typing, or pointing \cite{zadrozny2000natural}. Consequently, chatbots present potential as educational tools, particularly as information retrieval systems. By offering quick and convenient responses similar to human interaction, chatbots hold promise for facilitating research and education. This potential is evidenced by the rapid growth of OpenAI’s ChatGPT, an artificial intelligence chatbot that gained one million users within days of its launch \cite{Mortensen_2024}.



\subsection{Research Objectives}
\label{subsecsec:researchobjectives}

\subsubsection{General Objective}
\label{subsec:generalobjective}

The researchers aim to further expand the original digital ontology by \citeA{dimzon2015}, and develop a chatbot equipped with the ontology-based framework to answer questions about Panayanon folk narratives. Ultimately, the project output should be able to contribute to the preservation, accessibility and study of Panayanon folk literature.

\subsubsection{Specific Objectives}
\label{subsec:specificobjectives}
Specifically, the researchers aim to:
\begin{enumerate}
    \item Enhance the existing ontology by adding story elements as new classes, such as events and settings. Through this, additional details of the new folk narratives can be captured and queried.
    \item Expand the scope of the existing ontology by adding new entities, attributes, and relationships from Panayanon myths, legends, and folk tales. 
    \item Develop a prototype chatbot capable of understanding English questions and responding with accurate and appropriate information from the enhanced and expanded digital ontology.
\end{enumerate}


\subsection{Scope and Limitations of the Research}
\label{sec:scopelimitations}

The primary focus of this project is on the expansion and enhancement of the original digital ontology, which was first developed by \citeA{dimzon2015}. The scope of the folk literature analyzed for the digital ontology will be limited to folk narratives originating from the island of Panay, specifically myths, legends, and folk tales only. Further, these stories will be limited to those available during the project timeline, relying on existing research, expert consultations, and accessible resources. By building upon and expanding the original ontology, the researchers will ensure comprehensive coverage of the key entities and relationships within Panayanon folk narratives. Due to possible legal issues with ownership in data, the ontology will not contain the whole story but rather specific data that can help in research.

The native languages used in Panayanon folk narratives are Panayanon languages, namely Hiligaynon, Aklanon, and Karay-a. However, the language used in the development of the ontology and the chatbot will be in English. This is to ensure ease of use in academic and global contexts, thereby improving the accessibility of the ontology to a broader audience. Character names and other proper nouns will be kept in the original language to preserve authenticity.
To enhance the scope of the ontology, new classes will be created, such as events and settings, which were not present in the original ontology. The researchers will be consulting with literature experts to ensure that the new classes are relevant. As such, more classes may be introduced based on the suggestions of experts.

The chat bot will primarily be used as the tool for information retrieval from the ontology.  It will be developed as a prototype, focusing on demonstrating feasibility rather than full-scale deployment. Multilingual capabilities, advanced natural language processing for more complex queries, and deployment-level optimizations are beyond the scope of this project. Future projects may address these limitations.

\subsection{Significance of the Research}
\label{sec:significance}
The study holds significant value for the field of Panayanon cultural heritage and preservation for the following reasons:

The proposed system addresses the problem identified by \citeA{eugenio2007philippine} regarding the lack of published collections of Philippine folk literature. By serving as a central repository of knowledge for Panayanon folk narratives, the system is expected to facilitate easier access to Panayanon folk literature for researchers, students, educators, and the general public.

Additionally, the system seeks to address the issue of the decline of Panayanon oral literature by systematically collecting and digitizing these oral traditions, thereby contributing to their preservation for future generations.